---
title: "Lab 1: R review"
author: "Sophie Xu"
format: pdf
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
)
knitr::opts_chunk$set(eval = FALSE)
```

## Objects

1.  Create a code chunk and set the header parameter to TRUE and print out the top rows of the table with `head()` as above.

```{r load_data1}
data <- read.table(file = 
          url("http://www2.stat.duke.edu/~pdh10/FCBS/Exercises/azdiabetes.dat"), 
          header = TRUE)

head(data)
```

2.  Generate a sequence of $100$ equispaced real numbers from $0$ to $1$ and store it in a variable called `seq2`.

```{r seq}
seq2 <- seq(from = 0, to = 1, by = 0.01)
seq2
```

3.  Sort the entries in `seq3` from greatest to least.

```{r vector functions}
seq3 <- seq(from = -3, to = 3, by = .5)
sort(seq3, decreasing = TRUE)
```

## Matrices

4.  Find the variance of each row of `mat5`

```{r apply}
# generate large matrix
mat5 <- matrix(seq(1,100,1), nrow = 4, ncol = 25, byrow = T)


# apply(X = object, MARGIN = 1 for rows or 2 for columns, 
# FUN = function of choice)
# find variance of each row
apply(X = mat5, MARGIN = 1, FUN = var)
```

## Distribution Functions

5.  Generate $500$ samples from a Beta distribution with shape parameter $[a, b] = [0.5, 0.5]$ and store the samples in a variable called `W`

```{r beta}
W <- rbeta(500, shape1 = 0.5, shape2 = 0.5)
```

## Plotting

6.  Use code from above to make a few plots of your own.

```{r plot}
hist(W, 
     main = "Histogram of W", 
     xlab = "W", col = "lightblue")

plot(density(W), 
     main = "Density Plot of W", 
     xlab = "W")

boxplot(W, 
        main = "Boxplot of W", 
        ylab = "W")
```
