---
title: "Lab 4: Normal-Normal Conjugacy, Posterior Predictive Distribution and Monte Carlo"
output:
  pdf_document:
    toc: false
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)

options(
  htmltools.dir.version = FALSE,
  show.signif.stars = FALSE,
  digits = 3
)

knitr::opts_chunk$set(eval = FALSE)
```

## Instructions

Answer each question in the space provided by your instructor. Show your calculations and include R code where requested.

# 1. Normal-normal posterior for coffee-bag fill weights

A coffee-roasting company fills bags labeled as containing 500 grams. An inspector measures the fill weight of 12 bags:

```{r}
y <- c(503, 498, 501, 505, 499, 502,
       500, 504, 497, 501, 503, 499)

n <- length(y)
ybar <- mean(y)
```

Let \(\theta\) be the unknown mean fill weight. Assume

\[
Y_i\mid\theta\overset{\mathrm{ind}}{\sim}
\mathcal{N}(\theta,\sigma^2),
\qquad \sigma=4,
\]

and

\[
\theta\sim\mathcal{N}(m_0,\tau_0^2),
\qquad m_0=500,\qquad \tau_0=6.
\]

The normal-normal posterior is

\[
\theta\mid y\sim\mathcal{N}(m_n,v_n),
\]

where

\[
v_n=
\left(\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}\right)^{-1},
\qquad
m_n=
v_n\left(\frac{n\bar y}{\sigma^2}
+\frac{m_0}{\tau_0^2}\right).
\]

## Exercise 1: Posterior calculations

### Question 1

Calculate \(n\) and \(\bar y\).


### Question 2

Calculate the data precision, prior precision, posterior precision, \(m_n\), and \(v_n\).


### Question 3

Write the complete posterior distribution of \(\theta\mid y\).


### Question 4

Calculate the posterior mean, posterior variance, posterior standard deviation, and central 95\% credible interval.


### Question 5

Calculate

\[
\Pr(\theta>502\mid y).
\]


## Exercise 2: Density, CDF, and quantiles

### Question 1

Calculate the posterior density evaluated at 502 grams:

\[
f_{\theta\mid y}(502).
\]

Explain why this density value is not the probability that \(\theta=502\).


### Question 2

Calculate the posterior CDF at 502 grams:

\[
\Pr(\theta\leq502\mid y).
\]


### Question 3

Calculate

\[
\Pr(499<\theta<502\mid y)
\]

using two CDF values.


### Question 4

Use \(\texttt{qnorm()}\) to calculate the 2.5th percentile, 50th percentile, and 97.5th percentile of the posterior.


### Question 5

Verify that the 2.5th and 97.5th percentiles agree with the endpoints of the central 95\% credible interval.


# 2. Monte Carlo estimation and Bayesian decisions

For posterior draws \(\theta^{(1)},\ldots,\theta^{(M)}\),

\[
\mathbb{E}[g(\theta)\mid y]
\approx
\frac{1}{M}\sum_{m=1}^M g\left(\theta^{(m)}\right).
\]

Generate posterior draws:

```{r}
set.seed(402)

M <- 10000

theta_draws <- rnorm(
  M,
  mean = m_n,
  sd = sqrt(v_n)
)
```

## Exercise 3: Monte Carlo estimate of \(g(\theta)\)

### Question 1

Define

\[
g(\theta)=(\theta-500)^2
\]

and calculate the Monte Carlo estimator of

\[
\mathbb{E}[g(\theta)\mid y].
\]


### Question 2

Use the posterior draws to estimate \(\mathbb{E}(\theta\mid y)\) and \(\Pr(\theta>502\mid y)\).


## Exercise 4: Loss functions and Bayesian estimators

For this exercise, use

\[
\phi\mid y\sim\operatorname{Beta}(1,3).
\]

Under absolute error loss,

\[
L_1(a,\phi)=|a-\phi|.
\]

Under squared error loss,

\[
L_2(a,\phi)=(a-\phi)^2.
\]

The posterior expected loss of a candidate \(a\) is

\[
R(a)=\mathbb{E}[L(a,\phi)\mid y].
\]

In practice, posterior draws allow us to estimate this expected loss by

\[
\widehat{R}(a)
=
\frac{1}{M}\sum_{m=1}^M L(a,\phi^{(m)}).
\]

The posterior is strongly right-skewed, so its mean and median are visibly different. This lets us see the difference between squared-error and absolute-error decisions. For a normal posterior, the posterior mean and median coincide.

### Question 1

Generate 10,000 posterior draws and calculate the posterior mean and posterior median.


### Question 2

For candidate estimates \(a\in[0,1]\), calculate the empirical posterior expected absolute loss and squared loss:

\[
\widehat{R}_1(a)
=
\frac{1}{M}\sum_{m=1}^M|a-\phi^{(m)}|,
\]

\[
\widehat{R}_2(a)
=
\frac{1}{M}\sum_{m=1}^M(a-\phi^{(m)})^2.
\]


### Question 3

First plot the pointwise absolute loss

\[
L_1(a,\phi_0)=|a-\phi_0|
\]

for \(\phi_0=0.25\). Then plot the posterior expected absolute-loss curve and add vertical lines at the posterior mean and posterior median.


### Question 4

Find the value of \(a\) that minimizes the empirical absolute posterior loss and compare the absolute loss at the posterior mean and posterior median.


### Question 5

Find the value of \(a\) that minimizes the empirical squared posterior loss. Verify that it is close to the posterior mean.


# 3. Posterior predictive distribution

Let \(\widetilde Y\) be the fill weight of one new coffee bag. Conditional on \(\theta\),

\[
\widetilde Y\mid\theta\sim\mathcal{N}(\theta,\sigma^2).
\]

The posterior predictive distribution is

\[
p(\widetilde y\mid y)
=
\int
p(\widetilde y\mid\theta)
p(\theta\mid y)
\,d\theta.
\]

Because

\[
\widetilde Y=\theta+\varepsilon,
\qquad
\theta\mid y\sim\mathcal{N}(m_n,v_n),
\qquad
\varepsilon\sim\mathcal{N}(0,\sigma^2),
\]

we have

\[
\widetilde Y\mid y
\sim
\mathcal{N}(m_n,v_n+\sigma^2).
\]

# 4. Monte Carlo from the posterior predictive distribution

The two-stage simulation is

\[
\theta^{(m)}\sim p(\theta\mid y),
\qquad
\widetilde Y^{(m)}
\mid\theta^{(m)}
\sim\mathcal{N}(\theta^{(m)},\sigma^2).
\]

In R:

```{r}
set.seed(402)

theta_draws <- rnorm(
  M,
  mean = m_n,
  sd = sqrt(v_n)
)

y_new_draws <- rnorm(
  M,
  mean = theta_draws,
  sd = sigma
)
```

## Exercise 5: Analytic posterior predictive calculations

### Question 1

State the complete posterior predictive distribution of \(\widetilde Y\mid y\).


### Question 2

Calculate the posterior predictive mean, variance, and standard deviation.


### Question 3

Calculate a central 95% posterior predictive interval.


### Question 4

Calculate

\[
\Pr(\widetilde Y>505\mid y).
\]


### Question 5

Calculate the predictive quantiles at probabilities \(0.025\), \(0.50\), and \(0.975\).


## Exercise 6: Monte Carlo estimator from the posterior predictive distribution

### Question 1

Use the predictive draws to estimate the predictive mean, variance, standard deviation, probability above 505 grams, and central 95% predictive interval.


### Question 2

Distinguish between estimating \(\mathbb{E}[g(\theta)\mid y]\) using posterior draws and estimating \(\mathbb{E}[h(\widetilde Y)\mid y]\) using posterior predictive draws.


## Exercise 7: A general function of a future observation

Define

\[
h(\widetilde y)=\mathbb{I}(\widetilde y>505).
\]

### Question 1

Define \(h\) in R and calculate the Monte Carlo estimator of

\[
\mathbb{E}[h(\widetilde Y)\mid y].
\]


### Question 2

Explain why this Monte Carlo estimator is also an estimate of the posterior predictive probability \(\Pr(\widetilde Y>505\mid y)\).


# Useful R functions

| Task | R function |
|:---|:---|
| Draw from a normal distribution | \texttt{rnorm()} |
| Draw from a beta distribution | \texttt{rbeta()} |
| Normal density | \texttt{dnorm()} |
| Normal CDF | \texttt{pnorm()} |
| Normal quantile | \texttt{qnorm()} |
| Sample mean | \texttt{mean()} |
| Sample variance | \texttt{var()} |
| Sample standard deviation | \texttt{sd()} |
| Empirical quantiles | \texttt{quantile()} |
| Plot a curve | \texttt{plot()} |
| Add a line to a plot | \texttt{abline()} |
| Apply a function repeatedly | \texttt{sapply()} |

\vspace{1em}

The lab is made by Suchismita Roy. For questions, please email:
\texttt{suchismita.roy@duke.edu}.
