---
title: "Lab 4: Normal-Normal Conjugacy, Posterior Predictive Distribution and Monte Carlo"
author: Sophie Xu
format: pdf
execute:
  echo: true
  output: false
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)

options(
  htmltools.dir.version = FALSE,
  show.signif.stars = FALSE,
  digits = 3
)
```

# 1. Normal-normal posterior for coffee-bag fill weights

A coffee-roasting company fills bags labeled as containing 500 grams. An inspector measures the fill weight of 12 bags:

```{r}
y <- c(503, 498, 501, 505, 499, 502,
       500, 504, 497, 501, 503, 499)

n <- length(y)
ybar <- mean(y)
```

Let $\theta$ be the unknown mean fill weight. Assume

$$
Y_i\mid\theta\overset{\mathrm{ind}}{\sim}
\mathcal{N}(\theta,\sigma^2),
\qquad \sigma=4,
$$

and

$$
\theta\sim\mathcal{N}(m_0,\tau_0^2),
\qquad m_0=500,\qquad \tau_0=6.
$$

The normal-normal posterior is

$$
\theta\mid y\sim\mathcal{N}(m_n,v_n),
$$

where

$$
v_n=
\left(\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}\right)^{-1},
\qquad
m_n=
v_n\left(\frac{n\bar y}{\sigma^2}
+\frac{m_0}{\tau_0^2}\right).
$$

## Exercise 1: Posterior calculations

### Question 1

Calculate $n$ and $\bar y$.

```{r}
#| output: true
cat("n =", n, "\n")
cat("ybar =", ybar)
```

### Question 2

Calculate the data precision, prior precision, posterior precision, $m_n$, and $v_n$.

$$
\begin{gathered}
\text{data precision} = \frac{n}{\sigma^2} = \frac{12}{16} = 0.75 \\
\text{prior precision} = \frac{1}{\tau_0^2} = \frac{1}{36} = 0.0278 \\
\text{posterior precision} = \frac{n}{\sigma^2} + \frac{1}{\tau_0^2} = 0.7778 \\
v_n = \left(\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}\right)^{-1} = (0.7778)^{-1} = 1.2857 \\
\qquad
m_n = v_n\left(\frac{n\bar y}{\sigma^2}
+\frac{m_0}{\tau_0^2}\right) = 1.2857\left(\frac{12(501)}{16} + \frac{500}{36}\right) = 500.964
\end{gathered}
$$

### Question 3

Write the complete posterior distribution of $\theta\mid y$.

$$
\theta\mid y\sim\mathcal{N}(m_n,v_n) = \mathcal{N}(500.964, 1.2857)
$$

### Question 4

Calculate the posterior mean, posterior variance, posterior standard deviation, and central 95% credible interval.

$$
\begin{gathered}
\text{posterior mean} = m_n = 500.964 \\
\text{posterior variance} = v_n = 1.2857 \\
\text{posterior SD} = \sqrt{v_n} = 1.1339
\end{gathered}
$$

$$
\begin{aligned}
95\% \text{ Credible Interval} &= m_n \pm 1.96\sqrt{v_n} \\
&= 500.964 \pm 1.96(1.1339) \\
&= 500.964 \pm 2.2224 \\
&= (498.742,\ 503.187)
\end{aligned}
$$

### Question 5

Calculate $\Pr(\theta>502\mid y)$.

```{r}
#| output: true
1 - pnorm(502, mean = 500.964, sd = sqrt(1.2857))
```

So $\Pr(\theta>502\mid y) = 0.18$.

## Exercise 2: Density, CDF, and quantiles

### Question 1

To calculate the posterior density evaluated at 502 grams, I use the mean and variance I found before and use the normal distribution function in R:

```{r}
#| output: true
m_n <- 500.964
v_n <- 1.2857
sigma <- 4
dnorm(502, mean = m_n, sd = sqrt(v_n))
```

$$
f_{\theta\mid y}(502) = 0.23
$$

This density value is not the probability that $\theta=502$, because density values are just not probabilities. It comes from integrating over an interval. The $Pr(\theta=502) = 0$ because it is just a point on the pdf.

### Question 2

Calculate the posterior CDF at 502 grams:

```{r}
#| output: true
pnorm(502, mean = m_n, sd = sqrt(v_n))
```

$$
\Pr(\theta\leq502\mid y) = 0.82
$$

### Question 3

Calculate

$$
\Pr(499<\theta<502\mid y)
$$

using two CDF values.

$$
\Pr(499<\theta<502\mid y) = F_{\theta\mid y} (502) - F_{\theta\mid y} (499)
$$

```{r}
#| output: true
pnorm(502, mean = m_n, sd = sqrt(v_n)) - pnorm(499, mean = m_n, sd = sqrt(v_n))
```

And so $\Pr(499<\theta<502\mid y) = 0.78$.

### Question 4

Use $\texttt{qnorm()}$ to calculate the 2.5th percentile, 50th percentile, and 97.5th percentile of the posterior.

```{r}
#| output: true
q <- qnorm(c(0.025, 0.5, 0.975), mean = m_n, sd = sqrt(v_n))
cat("2.5th percentile =", q[1], "\n")
cat("50 th percentile =", q[2], "\n")
cat("97.5th percentile =", q[3], "\n")
```

### Question 5

The 2.5th and 97.5th percentiles here agree with the endpoints of the central 95% credible interval from Exercise 1 Q4: $(498.742,\ 503.187)$.

# 2. Monte Carlo estimation and Bayesian decisions

For posterior draws $\theta^{(1)},\ldots,\theta^{(M)}$,

$$
\mathbb{E}[g(\theta)\mid y]
\approx
\frac{1}{M}\sum_{m=1}^M g\left(\theta^{(m)}\right).
$$

Generate posterior draws:

```{r}
set.seed(402)

M <- 10000

theta_draws <- rnorm(M, mean = m_n, sd = sqrt(v_n))
```

## Exercise 3: Monte Carlo estimate of $g(\theta)$

### Question 1

To define

$$
g(\theta)=(\theta-500)^2
$$

It is the squared deviation of $\theta$ from 500.

And to calculate the Monte Carlo estimator of

$$
\mathbb{E}[g(\theta)\mid y] \\
$$

$$
\begin{aligned}
&= \frac{1}{M}\sum_{m=1}^{M}
g\left(\theta^{(m)}\right) \\
&= \frac{1}{M}\sum_{m=1}^{M}\left(\theta^{(m)} 
- 500\right)^2
\end{aligned}
$$

```{r}
#| output: true
mean((theta_draws - 500)^2)
```

$\approx 2.201$.

### Question 2

Use the posterior draws to estimate $\mathbb{E}(\theta\mid y)$ and $\Pr(\theta>502\mid y)$.

```{r}
#| output: true
cat("E(theta | y) =", mean(theta_draws), "\n")
cat("Pr(theta > 502 | y) =", mean(theta_draws > 502), "\n")
```

## Exercise 4: Loss functions and Bayesian estimators

For this exercise, use

$$
\phi\mid y\sim\operatorname{Beta}(1,3).
$$

Under absolute error loss,

$$
L_1(a,\phi)=|a-\phi|.
$$

Under squared error loss,

$$
L_2(a,\phi)=(a-\phi)^2.
$$

The posterior expected loss of a candidate $a$ is

$$
R(a)=\mathbb{E}[L(a,\phi)\mid y].
$$

In practice, posterior draws allow us to estimate this expected loss by

$$
\widehat{R}(a)
=
\frac{1}{M}\sum_{m=1}^M L(a,\phi^{(m)}).
$$

The posterior is strongly right-skewed, so its mean and median are visibly different. This lets us see the difference between squared-error and absolute-error decisions. For a normal posterior, the posterior mean and median coincide.

### Question 1

Generate 10,000 posterior draws and calculate the posterior mean and posterior median.

```{r}
#| output: true
M <- 10000
post <- rbeta(M, shape1 = 1, shape2 = 3)
cat("posterior mean =", mean(post), "\n")
cat("posterior median =", median(post), "\n")
```

### Question 2

For candidate estimates $a\in[0,1]$, calculate the empirical posterior expected absolute loss and squared loss:

$$
\widehat{R}_1(a)
=
\frac{1}{M}\sum_{m=1}^M|a-\phi^{(m)}|,
$$

$$
\widehat{R}_2(a)
=
\frac{1}{M}\sum_{m=1}^M(a-\phi^{(m)})^2.
$$

```{r}
a_grid <- seq(0, 1, by = 0.001)

R1 <- sapply(a_grid, function(a) mean(abs(a - post)))
R2 <- sapply(a_grid, function(a) mean(abs(a - post)^2))
```

### Question 3

First plot the pointwise absolute loss

$$
L_1(a,\phi_0)=|a-\phi_0|
$$

for $\phi_0=0.25$. Then plot the posterior expected absolute-loss curve and add vertical lines at the posterior mean and posterior median.

```{r}
#| output: true
phi0 <- 0.25
L1 <- abs(a_grid - phi0)

plot(a_grid, L1, type = "l",
     xlab = "a",
     ylab = expression(L[1](a, phi[0])),
     main = "Pointwise Absolute Loss, phi0 = 0.25")
```

```{r}
#| output: true
plot(a_grid, R1, type = "l",
     xlab = "a",
     ylab = expression(hat(R)[1](a)),
     main = "Posterior Expected Absolute Loss")
abline(v = mean(post), col = "blue", lty = 2)
abline(v = median(post), col = "red", lty = 2)
legend("topright", legend = c("posterior mean", "posterior median"),
       col = c("blue", "red"), lty = 2)
```

### Question 4

Find the value of $a$ that minimizes the empirical absolute posterior loss and compare the absolute loss at the posterior mean and posterior median.

```{r}
#| output: true
a_min <- a_grid[which.min(R1)]
cat("a* (minimizer) =", a_min, "\n")
cat("R1(a*) =", min(R1), "\n")
cat("R1(post_mean) =", mean(abs(mean(post) - post)), "\n")
cat("R1(post_median) =", mean(abs(median(post) - post)), "\n")
```

Here, we see that the value of $a$ that minimizes the empirical absolute posterior loss is 0.207, and R1(a\*) and R1(post_median) both give 0.155. The posterior median is the minimizer. The posterior mean gives a higher expected absolute loss.

### Question 5

Find the value of $a$ that minimizes the empirical squared posterior loss. Verify that it is close to the posterior mean.

```{r}
#| output: true

a_min2 <- a_grid[which.min(R2)]
cat("a* (minimizer) =", a_min2, "\n")
cat("R2(a*) =", min(R2), "\n")
cat("R2(post_mean) =", mean((mean(post) - post)^2), "\n")
```

Yep! It is close (basically the same) as the posterior mean.

# 3. Posterior predictive distribution

Let $\widetilde Y$ be the fill weight of one new coffee bag. Conditional on $\theta$,

$$
\widetilde Y\mid\theta\sim\mathcal{N}(\theta,\sigma^2),
\qquad
\theta\mid y\sim\mathcal{N}(m_n,v_n),
$$

and

$$
\widetilde Y\mid y
\sim
\mathcal{N}(m_n,v_n+\sigma^2).
$$

# 4. Monte Carlo from the posterior predictive distribution

The two-stage simulation is

$$
\theta^{(m)}\sim p(\theta\mid y),
\qquad
\widetilde Y^{(m)}
\mid\theta^{(m)}
\sim\mathcal{N}(\theta^{(m)},\sigma^2).
$$

In R:

```{r}
set.seed(402)

theta_draws <- rnorm(M, mean = m_n, sd = sqrt(v_n))

y_new_draws <- rnorm(M, mean = theta_draws, sd = sigma)
```

## Exercise 5: Analytic posterior predictive calculations

### Question 1

The complete posterior predictive distribution of $\widetilde Y\mid y$:

$$
\widetilde Y \mid y \sim \mathcal{N}(500.964,\ 17.2857)
$$

### Question 2

The posterior predictive mean, variance, and standard deviation:

$$
\begin{gathered}
\mathbb{E}(\widetilde Y\mid y) = m_n = 500.964 \\\text{Var}(\widetilde Y\mid y) = v_n + \sigma^2 = 1.2857 + 16 = 17.2857 \\SD = \sqrt{v_n+\sigma^2} = \sqrt{17.2857} \approx 4.1576
\end{gathered}
$$

### Question 3

A central 95% posterior predictive interval:

$$
\begin{aligned}
95\% \text{ Predictive Interval} &= m_n \pm 1.96\sqrt{v_n+\sigma^2} \\
&= 500.964 \pm 1.96(4.1576) \\
&= 500.964 \pm 8.1489 \\
&= (492.815,\ 509.113)
\end{aligned}
$$

### Question 4

Calculate

$$
\begin{aligned}
\Pr(\widetilde Y>505\mid y)
\end{aligned}
$$

```{r}
1 - pnorm(505, mean = m_n, sd = sqrt(v_n + sigma^2))
```

So $\Pr(\widetilde Y>505\mid y) = 0.166$.

### Question 5

Calculate the predictive quantiles at probabilities $0.025$, $0.50$, and $0.975$.

```{r}
#| output: true
q1 <- qnorm(c(0.025, 0.50, 0.975), mean = m_n, sd = sqrt(v_n + sigma^2))
cat("2.5th percentile =", q1[1], "\n")
cat("50 th percentile =", q1[2], "\n")
cat("97.5th percentile =", q1[3], "\n")
```

## Exercise 6: Monte Carlo estimator from the posterior predictive distribution

### Question 1

Use the predictive draws to estimate the predictive mean, variance, standard deviation, probability above 505 grams, and central 95% predictive interval.

```{r}
#| output: true
cat("predictive mean =", mean(y_new_draws), "\n")
cat("predictive variance =", var(y_new_draws), "\n")
cat("predictive SD =", sd(y_new_draws), "\n")
cat("Pr(Ynew > 505) =", mean(y_new_draws > 505), "\n")
cat("95% predictive interval = (", quantile(y_new_draws, c(0.025, 0.975)), ")")
```

### Question 2

To distinguish between estimating $\mathbb{E}[g(\theta)\mid y]$ using posterior draws and estimating $\mathbb{E}[h(\widetilde Y)\mid y]$ using posterior predictive draws, we know $\theta ^{(m)}$ draws only has uncertainly about the unknown mean $\theta$ after seeing the data; $\widetilde Y^{(m)}$ draws carries two layers of uncertainty, including parameter uncertainty and the noise $\epsilon$. So in code, `y_new_draws` uses `theta_draws` as an argument.

## Exercise 7: A general function of a future observation

Define

$$ h(\widetilde y)=\mathbb{I}(\widetilde y>505). $$

### Question 1

Define $h$ in R and calculate the Monte Carlo estimator of

$$ \mathbb{E}[h(\widetilde Y)\mid y]. $$

```{r}
h <- function(y) y >505
```

### Question 2

Explain why this Monte Carlo estimator is also an estimate of the posterior predictive probability $\Pr(\widetilde Y>505\mid y)$.
