---
title: "Lab 4: Normal-Normal Conjugacy, Posterior Predictive Distribution and Monte Carlo"
author: Sophie Xu
format: pdf
execute:
  echo: true
  output: false
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)

options(
  htmltools.dir.version = FALSE,
  show.signif.stars = FALSE,
  digits = 3
)
```

# 1. Normal-normal posterior for coffee-bag fill weights

A coffee-roasting company fills bags labeled as containing 500 grams. An inspector measures the fill weight of 12 bags:

```{r}
y <- c(503, 498, 501, 505, 499, 502,
       500, 504, 497, 501, 503, 499)

n <- length(y)
ybar <- mean(y)
```

Let $\theta$ be the unknown mean fill weight. Assume

$$
Y_i\mid\theta\overset{\mathrm{ind}}{\sim}
\mathcal{N}(\theta,\sigma^2),
\qquad \sigma=4,
$$

and

$$
\theta\sim\mathcal{N}(m_0,\tau_0^2),
\qquad m_0=500,\qquad \tau_0=6.
$$

The normal-normal posterior is

$$
\theta\mid y\sim\mathcal{N}(m_n,v_n),
$$

where

$$
v_n=
\left(\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}\right)^{-1},
\qquad
m_n=
v_n\left(\frac{n\bar y}{\sigma^2}
+\frac{m_0}{\tau_0^2}\right).
$$

## Exercise 1: Posterior calculations

### Question 1

Calculate $n$ and $\bar y$.

```{r}
#| output: true
cat("n =", n)
cat("ybar =", ybar)
```

### Question 2

Calculate the data precision, prior precision, posterior precision, $m_n$, and $v_n$.

$$
\begin{gathered}
\text{data precision} = \frac{n}{\sigma^2} = \frac{12}{16} = 0.75 \\
\text{prior precision} = \frac{1}{\tau_0^2} = \frac{1}{36} = 0.0278 \\
\text{posterior precision} = \frac{n}{\sigma^2} + \frac{1}{\tau_0^2} = 0.7778 \\
v_n = \left(\frac{n}{\sigma^2}+\frac{1}{\tau_0^2}\right)^{-1} = (0.7778)^{-1} = 1.2857 \\
\qquad
m_n = v_n\left(\frac{n\bar y}{\sigma^2}
+\frac{m_0}{\tau_0^2}\right) = 1.2857\left(\frac{12(501)}{16} + \frac{500}{36}\right) = 500.964
\end{gathered}
$$

### Question 3

Write the complete posterior distribution of $\theta\mid y$.

$$
\theta\mid y\sim\mathcal{N}(m_n,v_n) = \mathcal{N}(500.964, 1.2857)
$$

### Question 4

Calculate the posterior mean, posterior variance, posterior standard deviation, and central 95% credible interval.

$$
\begin{gathered}
\text{posterior mean} = m_n = 500.964 \\
\text{posterior variance} = v_n = 1.2857 \\
\text{posterior SD} = \sqrt{v_n} = 1.1339
\end{gathered}
$$

$$
\begin{aligned}
95\% \text{ Credible Interval} &= m_n \pm 1.96\sqrt{v_n} \\
&= 500.964 \pm 1.96(1.1339) \\
&= 500.964 \pm 2.2224 \\
&= (498.742,\ 503.187)
\end{aligned}
$$

### Question 5

Calculate

$$
\begin{aligned}
\Pr(\theta>502\mid y)
\end{aligned}
$$

```{r}
#| output: true
set.seed(222)
theta_sim <- rnorm(100000, mean = 500.964, sd = sqrt(1.2857))
mean(theta_sim > 502)
```
