---
title: "Lab 3: The Poisson model"
output:
  pdf_document:
    toc: false
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)

require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(bayesplot)
require(loo)
require(readxl)

options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
#knitr::opts_chunk$set(eval = FALSE)

ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(fig.align = 'center')
```

**Due:** 11:59 pm on Friday, February 13

## Turning in solutions

This lab is part of Homework 3. Solutions to the exercises, as well as the non-lab homework exercises are to be written up and uploaded to Gradescope as a PDF.

# Getting started

You will need the following R packages. If you do not already have them installed, please do so first using the `install.packages` function.

```{r eval = F}
require(tidyverse)
require(rstanarm)
require(magrittr)
require(rstan)
require(bayesplot)
require(loo)
require(readxl)
```

For this lab, you will need two stan files `lab-03-poisson-simple.stan` and `lab-03-poisson-simple.stan`, which you can download here:

- <https://omelikechi.github.io/sta402spring26/labs/lab03/lab-03-poisson-simple.stan>; and
- <https://omelikechi.github.io/sta402spring26/labs/lab03/lab-03-poisson-hurdle.stan>.

Download both and make sure to save them in the same folder as the R script or R markdown file you are working from.

# The Data

The data we have are counts of deaths per episode of Game of Thrones. Download the data (here: <https://omelikechi.github.io/sta402spring26/labs/lab03/GoT_Deaths.xlsx>) and save it locally to the same directory as your R markdown file. Once you have downloaded the data file into the SAME folder as your R markdown file, load the data by using the following R code.

```{r}
GoT <- read_xlsx("GoT_Deaths.xlsx", col_names = T)
head(GoT)
```

------------------------------------------------------------------------

1.  Plot a histogram of the death counts.

```{r, include=F}
ggplot(GoT, aes(x = Count))+
  geom_histogram(bins = 25, fill = "#deebf7") +
  labs(x = "Number of deaths", y = "Count")
```

------------------------------------------------------------------------

Because we have count data, it is natural to model the data with a Poisson distribution with parameter $\lambda$. If we take the empirical mean of the data as our estimate for $\lambda$, what does the sampling distribution look like?

```{r}
y <- GoT$Count
n <- length(y)

# Finish: obtain empirical mean
ybar <- mean(y)
sim_dat <- rpois(n, ybar)
qplot(sim_dat, bins = 20, xlab = "Simulated number of deaths", fill = I("#9ecae1"))

# create data frame for side-by-side histograms
df <- rbind(data.frame(y, "observed") %>% rename(count = 1, data = 2),
            data.frame(sim_dat, "simulated") %>% rename(count = 1, data = 2))
ggplot(df, aes(x = count, fill = data))+
  geom_histogram(position = "dodge", bins = 20) +
  scale_fill_brewer()  +
  labs(x = "Number of deaths", y = "Count")
```

Plotting the histogram of the data next to the histogram of the sampled data using $\hat{\lambda} = \bar{y}$, we see that there are many more 0-valued observations in the observed data than there are in the simulated data. We might suspect that the Poisson model will not be a good fit for the data. For now, let's proceed using the Poisson model and see how it behaves.

# Poisson model

We have learned that the Gamma distribution is conjugate for Poisson data. Here, we have chosen the prior $\lambda \sim Gamma(10,2)$ based on prior knowledge that Game of Thrones is a show filled with death. We run the model in Stan. We save the posterior sampled values of $\lambda$ in the variable `lambda_draws`.

```{r basic_mod,cache=T,message=F,warning=F}
stan_dat <- list(y = y, N = n)
fit <- stan("lab-03-poisson-simple.stan", data = stan_dat, refresh = 0, chains = 2)
lambda_draws <- as.matrix(fit, pars = "lambda")
```

------------------------------------------------------------------------

2.  Use the function `mcmc_areas()` to plot the smoothed posterior distribution for $\lambda$ with a 90% Highest Posterior Density region. Changing the `prob` parameter in `mcmc_areas()` will change the amount of probability mass that is highlighted. The highlighted region will begin in regions of highest density, and will move towards regions of lower density as `prob` gets larger.

```{r eval=F}
mcmc_areas(lambda_draws, prob = 0.9)
print(fit, pars = "lambda")
```

The line in the plot represents the posterior mean, and the shaded area represents 90% of the posterior density. How does the posterior mean for $\lambda$ compare to the sample mean?

------------------------------------------------------------------------
