---
title: "lab3"
author: Sophie Xu
format: pdf
execute:
  echo: true
  output: false
---

## The Data

```{r include=FALSE}
require(tidyverse) 
require(rstanarm) 
require(magrittr) 
require(rstan) 
require(bayesplot) 
require(loo) 
require(readxl)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )

ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(fig.align = 'center')
```

The data we have are counts of deaths per episode of Game of Thrones.

```{r read-data}
GoT <- read_xlsx("GoT_Deaths.xlsx", col_names = T)
head(GoT)
```

1.  Plot a histogram of the death counts.

```{r histogram-death-count}

hist(GoT$Count,
     main = "Histogram of Death Counts Per Episode",
     col = "skyblue",
     border = "white",
     xlab = "Death Counts")
```

The sampling distribution looks like this, centered around 3 to 4 deaths, roughly bell-shaped.

```{r sampling-distribution-poisson}
y <- GoT$Count
n <- length(y)
# Finish: obtain empirical mean
ybar <- mean(y)
sim_dat <- rpois(n, ybar)
qplot(sim_dat, bins = 20, xlab = "Simulated number of deaths", 
      fill = I("#9ecae1"))
```

Plotting the histogram of the data next to the histogram of the sampled data using $\hat{\lambda} = \bar{y}$, we see that there are many more 0-valued observations in the observed data than there are in the simulated data. We might suspect that the Poisson model will not be a good fit for the data. For now, let's proceed using the Poisson model and see how it behaves.

```{r histograms-side-by-side}
# create data frame for side-by-side histograms
df <- rbind(data.frame(y, "observed") %>% rename(count = 1, data = 2),
            data.frame(sim_dat, "simulated") %>% rename(count = 1, data = 2))
ggplot(df, aes(x = count, fill = data))+
  geom_histogram(position = "dodge", bins = 20) +
  scale_fill_brewer() +
  labs(x = "Number of deaths", y = "Count")
```

## Poisson Model

We have learned that the Gamma distribution is conjugate for Poisson data. Here, we have chosen the prior $\lambda \sim Gamma(10,2)$ based on prior knowledge that Game of Thrones is a show filled with death. We run the model in Stan. We save the posterior sampled values of $\lambda$ in the variable `lambda_draws`.

```{r basic_mod,cache=T,message=F,warning=F}
stan_dat <- list(y = y, N = n)
fit <- stan("lab-03-poisson-simple.stan", data = stan_dat, 
            refresh = 0, chains = 2)
lambda_draws <- as.matrix(fit, pars = "lambda")
```

2.  Use the function `mcmc_areas()` to plot the smoothed posterior distribution for $\lambda$ with a 90% Highest Posterior Density region. Changing the `prob` parameter in `mcmc_areas()` will change the amount of probability mass that is highlighted. The highlighted region will begin in regions of highest density, and will move towards regions of lower density as `prob` gets larger.

```{r eval=F}
mcmc_areas(lambda_draws, prob = 0.9)
print(fit, pars = "lambda")
```

The line in the plot represents the posterior mean, and the shaded area represents 90% of the posterior density. How does the posterior mean for λ compare to the sample mean?

The posterior mean for $\lambda$ is very similar to the sample mean, which is both very close to 4. This makes sense, since the posterior mean is a weighted average of the prior mean and the sample mean, and we only had $\beta = 2$ for the prior but 73 episodes for the data. So the posterior mean is dominated by the sample mean.
