---
title: "Lab 4: Normal-Normal Conjugacy, Posterior Predictive Distribution and Monte Carlo"
author: Sophie Xu
format: pdf
execute:
  echo: true
  output: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  results = "markup",
  message = FALSE,
  warning = FALSE,
  fig.width = 6.5,
  fig.height = 4.5,
  fig.align = "center"
)
```

# Part B. Computer analysis in R

**1. Load and inspect the data.**

```{r load-and-inspect}
library(gapminder)

d <- subset(gapminder, year == 2007)
d$log_gdp <- log(d$gdpPercap)
d$x <- d$log_gdp - mean(d$log_gdp)

summary(d[, c("lifeExp", "gdpPercap", "log_gdp")])
plot(d$log_gdp, d$lifeExp, pch = 19, col = "#2B78A6",
     xlab = "log(GDP per capita)", ylab = "Life expectancy (years)")
```

**2. Fit the conjugate model.** The next code block implements the normal--inverse-gamma posterior directly. We use weak but proper priors: $$
m_0=(65,0)^\mathsf{T},\qquad V_0=\mathrm{diag}(100,100),
\qquad a_0=2,\qquad b_0=100.
$$

```{r fit-conjugate-model}
y <- d$lifeExp
X <- cbind(1, d$x)
n <- nrow(X)

m0 <- c(65, 0)
V0 <- diag(c(100, 100))
a0 <- 2
b0 <- 100

Vn <- solve(solve(V0) + crossprod(X))
mn <- Vn %*% (solve(V0) %*% m0 + crossprod(X, y))
an <- a0 + n / 2
bn <- as.numeric(b0 + 0.5 * (
  crossprod(y) + t(m0) %*% solve(V0) %*% m0 -
  t(mn) %*% solve(Vn) %*% mn
))

mn
c(an = an, bn = bn)
```

**3. Draw from the joint posterior.** Here, `1/rgamma(...)` gives inverse-gamma draws under the convention used in class.

```{r joint-posterior}
set.seed(402)
S <- 10000
sigma2_draw <- 1 / rgamma(S, shape = an, rate = bn)

R <- chol(Vn)
beta_draw <- matrix(mn, nrow = S, ncol = 2, byrow = TRUE) +
  (matrix(rnorm(2 * S), nrow = S, ncol = 2) %*% R) *
  sqrt(sigma2_draw)

colnames(beta_draw) <- c("intercept", "slope")
apply(beta_draw, 2, quantile, probs = c(.025, .5, .975))
mean(beta_draw[, "slope"] > 0)
```

**4. Posterior and predictive summaries at \$20,000 GDP per capita.**

```{r posterior-and-predictive-summaries}
g_star <- 20000
x_star <- log(g_star) - mean(d$log_gdp)
x_star_vec <- c(1, x_star)

mu_star_draw <- as.vector(beta_draw %*% x_star_vec)
y_star_draw <- rnorm(S, mean = mu_star_draw, sd = sqrt(sigma2_draw))

quantile(mu_star_draw, c(.025, .5, .975))
quantile(y_star_draw, c(.025, .5, .975))
```

# Part C. What should you report?

1.  The 95% credible interval for the slope is (6.32, 8.07). This says that there is a 95% probability that for every 1 unit increase in log GDP per capita, the expected life expectancy increases by 6.32 - 8.07 years.
2.  Estimate $Pr(\beta_1>0\mid y)$ from the posterior draws. What does this posterior probability summarize?

```{r gtzero}
mean(beta_draw[, "slope"] > 0)
```

This posterior probability summarizes how certain we are that the slope is positive, which, in our case, exceeds 0.9999.

1.  Report a 95% credible interval for the conditional mean life expectancy at GDP per capita equal to \$20,000. Then report a 95% predictive interval for one new country's life expectancy at the same GDP level. Why are they different?
2.  Make a plot of the data with the posterior mean regression line:
